\section{Annex F: Normalization, Quantization, Depth Routing, and Runtime Algorithms}
\label{sec:annex-norm-bitnet-mod}

This annex collects the implementation-level detail for normalization variants, BitNet-style ternary quantization, adaptive depth routing (Mixture-of-Depths), conditional memory (Engram), factorized embeddings, and the runtime algorithms (training-step stability controls, SBERT inference router) that support the configuration-driven pipeline. The main sections defer to this annex for the mathematical formulations and pseudocode; the bibliography keys cited here resolve against \texttt{docs/bibliography/}.

\subsection{Normalization Variants: LayerNorm, RMSNorm, Dynamic Tanh, and Dynamic Erf}

Normalization determines how activation scale is controlled across depth. In this repository, normalization is not only a modeling choice but also a schema compatibility question, because only certain values are currently accepted by \texttt{norm\_type}. The schema contract is:
\[
\texttt{norm\_type} \in \{\texttt{layer\_norm}, \texttt{dynamic\_tanh}, \texttt{derf}\}
\]
so \texttt{rms\_norm} is \textbf{not} a valid schema value in the current contract. The three formulations most relevant to this codebase are:

\subsubsection{LayerNorm (Baseline)}
LayerNorm \citep{ba2016layer} is the default normalizer; it centers and rescales each token by its per-token mean and variance:
\[
\mu = \frac{1}{d}\sum_{i=1}^{d} x_i, \quad \sigma^2 = \frac{1}{d}\sum_{i=1}^{d}(x_i - \mu)^2, \quad y_i = \gamma_i \frac{x_i - \mu}{\sqrt{\sigma^2 + \epsilon}} + \beta_i.
\]

\subsubsection{RMSNorm (Not Accepted by Schema)}
RMSNorm removes mean-centering and only rescales by root mean square magnitude \citep{zhang_root_2019}:
\[
\mathrm{RMS}(x)=\sqrt{\frac{1}{d}\sum_{i=1}^{d}x_i^2+\epsilon},\qquad
y_i=\gamma_i\frac{x_i}{\mathrm{RMS}(x)}.
\]
Compared with LayerNorm, RMSNorm is computationally simpler (no subtraction of feature mean) and is often used when reducing normalization overhead is important. It is documented here for reference but is \textbf{not} an accepted \texttt{norm\_type} value.

\subsubsection{Dynamic Tanh (DyT)}
Dynamic Tanh proposes replacing explicit normalization with a bounded elementwise map \citep{zhu_transformers_2025}:
\[
\mathrm{DyT}(x)=\tanh(\alpha x)
\]
where $\alpha$ is learned. The core idea is that bounded nonlinear contraction can provide stable signal scaling without explicitly computing per-token normalization statistics.

\subsubsection{Dynamic Erf (Derf)}
Derf extends the same normalization-free direction by using an error-function based map \citep{chen_stronger_2025}:
\[
\mathrm{Derf}(x)=\mathrm{erf}(\alpha x+s)
\]
with learnable scale/shift. Reported results in the cited work indicate stronger performance than DyT and common normalization baselines across multiple domains.

\subsubsection{Comparison}
\scriptsize
\begin{longtable}{p{2.2cm}p{3.4cm}p{2.0cm}p{4.0cm}}
\toprule
\textbf{Method} & \textbf{Formula} & \textbf{Stats Needed} & \textbf{Notes} \\
\midrule
\endhead
LayerNorm & $\gamma_i (x_i - \mu)/\sqrt{\sigma^2 + \epsilon} + \beta_i$ & mean + variance & Default; full centering and rescaling \citep{ba2016layer}. \\
RMSNorm & $\gamma_i x_i/\sqrt{\frac{1}{d}\sum_j x_j^2+\epsilon}$ & RMS only & Lower overhead; widely used baseline; \textbf{not accepted by schema} \citep{zhang_root_2019}. \\
Dynamic Tanh & $\tanh(\alpha x)$ & none & Normalization-free bounded transform; drop-in replacement \citep{zhu_transformers_2025}. \\
Dynamic Erf & $\mathrm{erf}(\alpha x+s)$ & none & Normalization-free alternative; improves over DyT \citep{chen_stronger_2025}. \\
\bottomrule
\end{longtable}
\normalsize

\subsection{BitNet Ternary Quantization Path}

When \texttt{use\_bitnet=true}, BitLinear layers replace standard \texttt{nn.Linear} with ternary weight quantization ($-1, 0, +1$) following BitNet \citep{wang_bitnet_2023}. Given weight tensor $W$, a practical scaling is:
\[
s=\text{mean}(|W|),\qquad
\tilde{W}=\text{clip}\left(\text{round}\left(\frac{W}{s}\right),-1,1\right).
\]
The packed mapping uses two bits per weight symbol for storage efficiency. This enables significant compression ($\sim 32\times$ versus FP32) but is experimental and may affect model quality; it is best suited for inference deployment.

Activation quantization for the INT8 path is:
\[
q=\text{round}\left(x\cdot\frac{127}{\max(|x|)+\epsilon}\right),\qquad q\in[-128,127]
\]
with dequantization $x\approx q/\alpha$. For $N$ parameters, size estimates before metadata and packing overhead are:
\[
\text{FP32 size}\approx 4N,\quad \text{FP16 size}\approx 2N,\quad \text{1.58-bit size}\approx \frac{1.58}{8}N
\]
which aligns with lightweight deployment goals \citep{samson_lightweight_2026,bravin_embbert_2026}.

\subsection{Mixture-of-Depths (MoD) Token Routing}

When \texttt{use\_mixture\_of\_depths=true}, each transformer block scores tokens with a lightweight router. Only the top \texttt{mixture\_of\_depths\_capacity\_ratio} subset is updated; skipped tokens pass through unchanged \citep{zhu2026mixtureofdepthsattention}. This allocates more depth to salient tokens, reducing per-layer compute.

\begin{algorithmic}[1]
\State \textbf{Input:} hidden states $H \in \mathbb{R}^{n \times d}$, router $R$, capacity ratio $\rho$
\State $\text{scores} \gets R(H)$ \Comment{per-token router score}
\State $\mathcal{S} \gets \texttt{top-k}(\text{scores}, k=\lceil \rho \cdot n \rceil)$ \Comment{select salient subset}
\State $H_{\text{update}} \gets \texttt{block}(H[\mathcal{S}])$ \Comment{apply transformer block to selected tokens}
\State $H[\mathcal{S}] \gets H_{\text{update}}$ \Comment{write back; non-selected tokens bypass the block}
\State \textbf{Output:} $H$ \Comment{auxiliary load-balancing loss added from \texttt{mixture\_of\_depths\_router\_aux\_loss\_weight}}
\end{algorithmic}

\subsubsection{Configuration}
\begin{itemize}[leftmargin=*]
    \item \texttt{use\_mixture\_of\_depths}: bool --- enable MoD token routing.
    \item \texttt{mixture\_of\_depths\_capacity\_ratio}: float in $(0, 1]$ --- fraction of tokens updated per layer.
    \item \texttt{mixture\_of\_depths\_router\_aux\_loss\_weight}: float $\ge 0$ --- auxiliary loss weight for router regularization.
\end{itemize}

\subsection{Engram Conditional Memory}

Engram (\texttt{engram\_attn}) implements conditional memory via scalable N-gram lookup \citep{cheng2026conditionalmemoryscalablelookup}. It builds hash-based lookup tables for N-gram contexts from size 2 up to \texttt{engram\_max\_ngram\_size}, with independent hash heads per N-gram order. A causal depthwise convolution (\texttt{engram\_kernel\_size}) provides short-range context before the N-gram lookup.

\begin{algorithmic}[1]
\State \textbf{Input:} token stream $x_{1:n}$, N-gram orders $\{2, \ldots, N\}$, hash heads per order $H$
\State $h_{\text{conv}} \gets \text{DepthwiseConv1d}(x_{1:n}, k=\texttt{engram\_kernel\_size})$ \Comment{short-range context}
\For{each N-gram order $k \in \{2, \ldots, N\}$}
    \State $\text{ctx}_t^{(k)} \gets x_{t-k+1:t}$ \Comment{N-gram context window}
    \State $\text{lookup}^{(k)}_t \gets \text{HashTable}^{(k)}[\text{hash}(\text{ctx}_t^{(k)})]$ \Comment{per-head retrieval, $H$ heads per order}
\EndFor
\State $y_t \gets \sum_{k} \text{lookup}^{(k)}_t + h_{\text{conv},t}$ \Comment{combine conditional memories}
\State \textbf{Output:} $y_{1:n}$
\end{algorithmic}

\subsubsection{Configuration}
\begin{itemize}[leftmargin=*]
    \item \texttt{engram\_max\_ngram\_size}: int $\ge 2$ --- highest N-gram order (default 3).
    \item \texttt{engram\_n\_heads\_per\_ngram}: int $\ge 1$ --- hash heads per N-gram order (default 4).
    \item \texttt{engram\_embed\_dim\_per\_head}: int $\ge 1$ --- embedding dim per hash head (default 32).
    \item \texttt{engram\_kernel\_size}: int $\ge 1$ --- causal depthwise conv kernel width (default 4).
    \item \texttt{engram\_seed}: int --- hash seed for reproducibility (default 42).
\end{itemize}
Total Engram hidden size = $(\texttt{max\_ngram\_size} - 1) \times \texttt{n\_heads\_per\_ngram} \times \texttt{embed\_dim\_per\_head}$.

\subsection{Factorized Embeddings and Embedding Convolution}

\subsubsection{Factorized Embeddings}
When \texttt{use\_factorized\_embedding=true}, the embedding matrix is factorized into two smaller matrices, reducing parameters from $O(V \times H)$ to $O(V \times R + R \times H)$ where $R = \texttt{factorized\_embedding\_dim}$.
\begin{itemize}[leftmargin=*]
    \item \texttt{use\_factorized\_embedding}: bool --- enable factorization.
    \item \texttt{factorized\_embedding\_dim}: int $\ge 1$ --- intermediate dimension (typical: 64--256).
\end{itemize}

\subsubsection{Embedding Convolution}
When \texttt{use\_embedding\_conv=true}, a depthwise Conv1d is applied over the embedding stream before the first transformer block, with kernel size \texttt{embedding\_conv\_kernel}. This provides a lightweight local-context filter on token embeddings.

\subsection{Training-Step Stability Controls}

The schema-driven training step applies gradient accumulation, global-norm clipping, post-clip explosion guards, and bounded NaN/Inf retries. The accumulated gradient is:
\[
g_{\text{acc}}=\frac{1}{K}\sum_{i=1}^{K} g_i,\quad K=\texttt{gradient\_accumulation\_steps}
\]
\[
g_{\text{clip}} = g_{\text{acc}}\cdot
\min\left(1,\frac{\tau}{\|g_{\text{acc}}\|_2+\epsilon}\right),\quad
\tau=\texttt{grad\_clip\_max\_norm}
\]
then overflow guards use \texttt{inf\_post\_clip\_threshold} and retry logic bounded by \texttt{max\_nan\_retries}.

\begin{algorithm}[t]
\caption{Schema-Driven Training Step with Stability Controls}
\begin{algorithmic}[1]
\Require Batch stream, config $C$
\State Initialize retry counter $r\gets 0$
\For{each optimizer step}
    \State Accumulate gradients for $K=C.\texttt{gradient\_accumulation\_steps}$ micro-batches
    \State Apply global norm clipping with $\tau=C.\texttt{grad\_clip\_max\_norm}$
    \If{post-clip gradient exceeds $C.\texttt{inf\_post\_clip\_threshold}$ or NaN/Inf detected}
        \If{$r < C.\texttt{max\_nan\_retries}$}
            \State restore safe state / skip step; $r \gets r+1$
            \State \textbf{continue}
        \Else
            \State stop training with failure state
        \EndIf
    \EndIf
    \State run optimizer step selected by \texttt{optimizer\_class}
    \State update scheduler (\texttt{cosine}, \texttt{constant}, or \texttt{linear\_warmup\_then\_constant})
    \If{step mod \texttt{checkpoint\_every\_n\_steps}$=0$}
        \State save rolling checkpoint and prune to \texttt{max\_rolling\_checkpoints}
    \EndIf
    \State update best checkpoints up to \texttt{num\_best\_checkpoints}
    \State emit CSV + telemetry following \texttt{gradient\_log\_interval} and \texttt{telemetry\_log\_interval}
\EndFor
\end{algorithmic}
\end{algorithm}

\subsection{SBERT Inference Mode Router}

The SBERT inference dispatcher selects among similarity scoring, corpus search, clustering, and persistent encoding modes over a single shared encoder \citep{reimers_sentence-bert_2019}. The cosine similarity and regression loss are:
\[
\text{cos}(e_1,e_2)=\frac{e_1^\top e_2}{\|e_1\|\|e_2\|},\qquad
\mathcal{L}_{\text{cos}}=\left(\text{cos}(e_1,e_2)-y\right)^2
\]
with $y\in[-1,1]$ in this pipeline.

\begin{algorithm}[t]
\caption{SBERT Inference Mode Router}
\begin{algorithmic}[1]
\Require mode $m$, model $E$, inputs $X$
\If{$m=\texttt{similarity}$}
    \State return $\cos(E(x_1),E(x_2))$
\ElsIf{$m=\texttt{search}$}
    \State return top-$k$ by dot-product/cosine against corpus embeddings
\ElsIf{$m=\texttt{cluster}$}
    \State return clustering labels over $E(X)$
\Else
    \State return serialized embeddings $E(X)$
\EndIf
\end{algorithmic}
\end{algorithm}